---
title: "Inferential Data Analysis of an Exponential Distribution"
author: "John Akwei, ECMp ERMp"
date: "Sunday, February 15, 2015"
output: html_document
---

#### Synopsis  
This document generates a simulation, (in the R programming language), of the distribution of the mean of 40 exponentials. The relationship of the two distributions is then verified with the Central Limit Theorem.  
```{r, echo=F, message=F, warning=F}
options(scipen=999)
if (!require("ggplot2")) { install.packages("ggplot2"); require("ggplot2") }
```

#### Simulation of an Exponential Distribution  
The following R progamming language code creates a distribution of averages of 40 exponentials, with rexp(n, lambda), where lambda is the rate parameter.  

The mean of exponential distribution is 1/lambda. Lambda is set at 0.2 for all of the simulations. A thousand simulation are performed, in order to increase the power of the analysis.  
```{r, echo=F}
# Parameters:
set.seed(7); lambda <- 0.2; m <- 40; n <- 1000
# Variables:
simulationArray <- replicate(m, rexp(n, lambda))
simulationMean <- apply(simulationArray, 2, mean)
# Graph:
hist(simulationMean, xlab="Mean", ylab="Frequency", main="Distribution of the Mean of 40 Exponentials", col="blue")
```

#### Comparison of the sample mean of the distribution to theoretical mean  
Sample mean:  
```{r, echo=F}
sampleMean <- mean(simulationMean); sampleMean
```
Theoretical mean:  
```{r, echo=F}
theoreticalMean <- 1/lambda; theoreticalMean
```

#### Conclusion:  
The analysis demonstrates the center of the distribution of the sample is approximately the theoretical center.  

#### Compararison of sample variance to theoretical variance  
Sample Deviation:  
```{r, echo=F}
sampleDeviation <- sd(simulationMean); sampleDeviation
```
Theoretical Deviation:  
```{r, echo=F}
theoreticalDeviation <- (1/lambda)/sqrt(n); theoreticalDeviation
```
Sample Variance:  
```{r, echo=F}
sampleVariance <- sampleDeviation^2; sampleVariance
```
Theoretical Variance:  
```{r, echo=F}
theoreticalVariance <- ((1/lambda)*(1/sqrt(n)))^2; theoreticalVariance
```

#### Conclusion:  
The standard variance is close to the theoretical variance.  

#### Graph: Normality of the exponential distribution  
```{r, echo=F, message=F, warning=F}
alignment <- function(align=c("left", "center", "right"))
{
    # get alignment choice
    align <- match.arg(align)

    left.margin <- c(left="0", center="auto", right="auto")
    right.margin <- c(left="auto", center="auto", right="0")

    sprintf("display:block; margin-left: %s ; margin-right: %s ;", left.margin[align], right.margin[align])
}
alignment("center")
plotdata <- data.frame(simulationMean)
ggplot(plotdata, aes(x=simulationMean)) +
geom_histogram(aes(y=..density..), col="blue", fill="blue") +
geom_density(col="orange", size=1) +
xlab("Mean") +
ylab("Density") +
ggtitle("Normality of the Distribution")
```

#### Graph: Verification of normality of the tests  
```{r, echo=F, message=F, warning=F}
alignment <- function(align=c("left", "center", "right"))
{
    # get alignment choice
    align <- match.arg(align)

    left.margin <- c(left="0", center="auto", right="auto")
    right.margin <- c(left="auto", center="auto", right="0")

    sprintf("display:block; margin-left: %s ; margin-right: %s ;", left.margin[align], right.margin[align])
}
alignment("center")
qqnorm(simulationMean, col="blue"); qqline(simulationMean, col="orange")
```

## Final Conclusion  
The Exponential Simulation in this document follows the characteristics of the Central Limit Theorem.  