## PA1_template.Rmd  

## Peer_Assessment1  

author: "John Akwei"  
date: "Wednesday, January 14, 2015"  
output: html_document  
application: "Reproducible Research: Peer Assessment 1"  

#### The data analyzed within this document was generated by a personal activity monitoring device, (Fitbit, Nike Fuelband, etc.), from an anonymous individual, collected during the months of October and November 2012, and includes the number of steps taken in 5 minute intervals each day.  

### 1)  
### Loading and Preprocessing the Data  

#### Required packages  
```{r}
required <- function(wd) {
  setwd(wd)
  if (!require("data.table")) { install.packages("data.table"); require("data.table") }
  if (!require("knitr")) { install.packages("knitr"); require("knitr") }
  if (!require("ggplot2")) { install.packages("ggplot2"); require("ggplot2") }
}
required("C:/Users/johnakwei/Desktop/Coursera/ReproducibleResearch/Week2/RepData_PeerAssessment1")
```

#### Data download and extraction  
```{r}
unextracted <- "activity.zip"
extracted  <- "activity"
dataLocation <- "http://d396qusza40orc.cloudfront.net/repdata/data/activity.zip"
if (!file.exists(unextracted)) { download.file(dataLocation, unextracted) }
if (!file.exists(extracted)) { unzip(unextracted) }
```

#### Preprocessing the data  
```{r}
dataFrame <- read.csv("activity.csv")
```

### Analysis of Personal Activity Monitoring Device Data:  

### 2)  
### What is the mean total number of steps taken per day?  
```{r}
dataAvg <- tapply(dataFrame$steps, dataFrame$date, FUN=sum, na.rm=T)
qplot(dataAvg, binwidth=1000, main="Total Step/Day", xlab="Interval", ylab="Total Number of Steps")
mean(dataAvg, na.rm=T)
median(dataAvg, na.rm=T)
```

### 3)  
### What is the average daily activity pattern?  
```{r}
dailyPattern <- aggregate(x=list(steps=dataFrame$steps), by=list(interval=dataFrame$interval), FUN=mean, na.rm=T)
```

```{r}
plot(dailyPattern, type="l", main="Daily Activity Pattern", xlab="5 Minute Intervals", ylab="Average Steps Taken", col="blue")
```

#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r}
dailyPattern[which.max(dailyPattern$steps),]
```

### 4)  
### Imputing missing values  

#### Total number of missing values in the dataset  
```{r}
naValues <- is.na(dataFrame$steps)
table(naValues)
```

#### Filled in missing values with 5-minute interval mean values  
```{r}
replaceValues <- function(steps, interval) {
  newValues <- NA
  if (!is.na(steps)) newValues <- c(steps)
  else newValues <- (dailyPattern[dailyPattern$interval==interval, "steps"])
  return(newValues) }
```

#### Dataset with missing data filled in  
```{r}
newValuesDF <- dataFrame
newValuesDF$steps <- mapply(replaceValues, newValuesDF$steps, newValuesDF$interval)
```

#### Histogram of the total steps/day  
```{r}
dataAvg <- tapply(newValuesDF$steps, newValuesDF$date, FUN=sum)
hist(dataAvg, main="Histogram of Steps/Day", xlab="Interval", ylab="Steps", col="green")
```

#### Mean/Median of the total steps/day  
```{r}
mean(dataAvg)
median(dataAvg)
```

### Question:  
#### Do these values differ from estimates from the first part of the assignment?  
#### What is the impact of imputing missing data on the estimates of the total daily number of steps?  

### Answer:  
#### The Mean/Median differs from the beginning estimates. Via creating a dataset with filled missing values, the resulting Mean/Median increased.  

### 5)  
### Are there differences in activity patterns between weekdays and weekends?  
```{r}
weekLabels <- function(date) {
  if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")) { "Weekend"}
  else { "Weekday" } }
newValuesDF$weekLabels <- as.factor(sapply(newValuesDF$date, weekLabels))
```

#### Plot of the 5-minute interval, and the average number of steps taken, averaged across all weekdays or weekend days  
```{r}
par(mfrow=c(2,1))
for (type in c("Weekend", "Weekday")) {
intervalAvg <- aggregate(steps~interval, data=newValuesDF,
subset=newValuesDF$weekLabels==type, FUN=mean); plot(intervalAvg, type="l", main=type, col="blue") }
```